---
title: "Getting Started with mosaicresid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mosaicresid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mosaicresid)
library(ggplot2)
```

## Why this package exists

Base R's `mosaicplot(shade = TRUE)` is a genuinely useful plot: tile area
encodes joint proportion, and tile color encodes how far each cell departs
from what independence would predict (a Pearson residual). But it's drawn
with base graphics, so it can't be faceted, themed, or composed with other
`ggplot2` layers, and it returns nothing you can inspect -- the numbers only
ever reach a device.

`mosaicresid` reproduces the same statistics and the same true mosaic
geometry (recursive partitioning, not a grid-of-equal-cells approximation),
but as an ordinary `ggplot2` layer, mapped the way every other `ggplot2`
geom is mapped:

```{r}
ggplot(mtcars) +
  geom_mosaic_residual(aes(x = factor(cyl), y = factor(am))) +
  theme_minimal() +
  labs(title = "Cylinders vs Transmission", x = NULL, y = NULL)
```

That single call did everything base R's `mosaicplot(shade = TRUE)` does:

- **Sized** each tile by its joint proportion (true mosaic geometry: width
  from the first variable's marginal split, height from the second
  variable's split *conditional* on the first).
- **Computed** a two-way independence model and the Pearson residual of each
  cell against it.
- **Colored** each tile by that residual, via a diverging red-white-blue
  scale.
- **Labeled** each tile in place, since the usual axis-tick labeling doesn't
  work for a mosaic plot's inner variable (its split positions differ from
  column to column -- there's no one shared axis position that's correct
  for the whole plot).

No manual `stat_*()` call, no manual residual computation, no manual scale.

## Working with raw or pre-aggregated data

`geom_mosaic_residual()` accepts either shape of input, through the same
code path. One row per observation:

```{r}
titanic_raw <- as.data.frame(Titanic)
titanic_raw <- titanic_raw[rep(seq_len(nrow(titanic_raw)), titanic_raw$Freq), ]

ggplot(titanic_raw) +
  geom_mosaic_residual(aes(x = Class, y = Survived)) +
  theme_minimal()
```

Or already-aggregated counts, via `weight`:

```{r}
titanic_counts <- as.data.frame(Titanic)
titanic_counts <- aggregate(Freq ~ Class + Survived, titanic_counts, sum)

ggplot(titanic_counts) +
  geom_mosaic_residual(aes(x = Class, y = Survived), weight = "Freq") +
  theme_minimal()
```

(`weight` is passed through `...` to `stat_mosaic_residual()`/`stat_mosaic_label()`;
it isn't a literal `geom_mosaic_residual()` argument, so it works the same way
`weight` works for any other `ggplot2` stat.)

## Overriding the color scale

The default scale is an ordinary `ggplot2` scale object, so it overrides
the usual way -- and labels keep working fine when you do:

```{r}
ggplot(mtcars) +
  geom_mosaic_residual(aes(x = factor(cyl), y = factor(am))) +
  scale_fill_gradient2(low = "purple", mid = "white", high = "darkgreen") +
  theme_minimal()
```

## Spacing, labels, and small cells

```{r}
ggplot(mtcars) +
  geom_mosaic_residual(
    aes(x = factor(cyl), y = factor(gear)),
    spacing = 0.02,
    min_label_width = 0.04,
    min_label_height = 0.06,
    label.size = 2.5
  ) +
  theme_minimal() +
  labs(title = "Cylinders vs Gears", x = NULL, y = NULL)
```

Cells smaller than `min_label_width`/`min_label_height` (in 0-1 fractional
plot units) are still colored, just left unlabeled, rather than overflowing
text across tile boundaries.

## Matching base R's `dir` and `type` options

Base R's `mosaicplot()` lets you force the split direction per variable
(`dir`) and choose among three residual formulas (`type`). Both are
supported here with the same semantics:

```{r}
ggplot(mtcars) +
  geom_mosaic_residual(
    aes(x = factor(cyl), y = factor(am)),
    dir = c("h", "v"),   # swap which variable splits vertically first
    type = "deviance"    # pearson (default), deviance, or FT (Freeman-Tukey)
  ) +
  theme_minimal()
```

## Inspecting the numbers directly

Because the statistics are computed by a plain function, not hidden inside
a device call, you can get the tile geometry and residuals as an ordinary
data frame -- useful for checking the math, or for driving a completely
custom plot:

```{r}
compute_mosaic_residuals(mtcars, "cyl", "am")
```

## What this package deliberately does not do

`compute_mosaic_residuals()`/`add_independence_residuals()` only support
two grouping variables. Past two variables, "expected count" stops having
one unambiguous definition -- complete independence, joint independence,
conditional independence, and all-two-way-associations are different,
non-equivalent loglinear models, and the caller has to pick one. Rather
than silently computing a plausible-looking but wrong number, calling these
functions with more than two variables fails loudly and points toward
`stats::loglin()`/`glm(family = poisson)` as the general-purpose tools for
fitting a model you've chosen explicitly. See `vignette("comparison")` for
how this scope decision compares to the alternatives.
